---
title: 'Lab#9'
author: "Ifeoma Okonye"
date: "2024-04-02"
output: html_document
---

## Introduction {.tabset}

For this lab, I used the raster, sf and tidyverse packages. The data layers I made use of are the 2019 Cropland Data Layer (CDL) for Riley county and elevation data. I did some data tidying by converting the 0 values in the land use data to NA and then I converted both dataset (CDL and elevation) to the right coordinate reference system.

```{r message=FALSE, warning=FALSE}
library(pacman)
p_load(raster, sf, tidyverse)


riley <- raster("CDL_2019_20161.tif")
elev <- raster("elevation.tif")


plot(riley)

riley[riley == 0] <- NA

plot(riley)

crs(riley)
crs(elev)

crs <- '+proj=utm +zone=11 +ellps=GRS80 +datum=NAD83 +units=m +no_defs'

riley <- projectRaster(riley, crs = crs, method = "ngb") 

elev <- projectRaster(elev, crs = crs, method = "bilinear")

plot(elev)
plot(riley)

```


### Question 1

I transformed both raster so that they have a consistent grid structure and aggregated them to a grid cell resolution of 1km x 1km. I then plotted the resampled raster layers.

```{r message=FALSE, warning=FALSE}

# for croptype 2019 riley data
new_ras_1km <- raster(crs = crs,
                  ext = extent(riley),
                  resolution = c(1000, 1000))


riley_1km <- resample(riley, new_ras_1km,
                                  method = "ngb")

plot(riley_1km, main = "2019 Riley Landuse Resampled to 1km")



# for elevation data

elevras_1km <- raster(crs = crs,
                  ext = extent(riley),
                  resolution = c(1000, 1000))


elev_1km <- resample(elev, elevras_1km,
                                  method = "bilinear")

plot(elev_1km, main = "Riley Elevation Resampled to 1km")


```




### Question 2

I ran the smooth focal statistics to the elevation values and ran the same focal statistics to the landuse dataset to determine local diversity in the data. I plotted the resulting rasters.

```{r message=FALSE, warning=FALSE}

# focal statistics on elevation data

elev_smooth <- focal(elev_1km, 
                     w = matrix(1 , ncol = 3, nrow = 3), 
                     fun = mean, na.rm = T)
plot(elev_smooth)


# focal statistics on landuse data

riley_smooth <- focal(riley_1km, 
                      w = matrix(1, ncol = 3, nrow = 3), 
                      fun = modal, na.rm = T)
plot(riley_smooth)  

```


### Question 3

Using the smoothed elevation raster, I calculated and reported the average slope and elevation for each land use type in the raster. The landuse type at the highest elevation is Sorghum (4) at an elevation of 409m and the landuse at the lowest elevation is Open Water (111) at an elevation 0f 339m. This analysis seems to be reasonable as Open Water will be found at lower elevations and crops like Sorghum are found at higher elevations.

For the slope analysis, the slope was was categorized into 4 zones with 4 being the highest slope zone and 1 being the lowest. From the analysis, it is difficult to identify which land use type fall within each zone.


```{r message=FALSE, warning=FALSE}

#fix extent
elev_smooth_rc <- crop(elev_smooth, extent(riley_1km))
plot(elev_smooth_rc)

#check resolution
res(elev_smooth)
res(riley_1km)

# zonal op
elev_crop <- zonal(elev_smooth_rc, riley_1km, 
                   fun = mean, na.rm=T) %>% 
  as.data.frame()
round(elev_crop)



#calculating slope (terrain)
slope <- terrain(elev_smooth_rc, 
                 opt = "slope", 
                 units = "degrees") 

plot(slope)
hist(slope)


cellStats(slope, stat = min)
cellStats(slope, stat = max)

#reclassifying slopes into 4 different zones
zones <- reclassify(slope, c(0.0002, 0.0075, 1,      # From 0.0002 to 0.0075 -> Zone 1
                              0.0075, 0.015, 2,      # From 0.0075 to 0.015 -> Zone 2
                              0.015, 0.0225, 3,      # From 0.015 to 0.0225 -> Zone 3
                               0.0225, 0.03, 4),     # From 0.0225 to 0.03 -> Zone 4
                            byrow = TRUE)

zones <- ratify(zones)
plot(zones)

#getting average slope for each landuse type

landuse_slope <- zonal(riley_1km, zones, fun = mean) 
landuse_slope

```


### Question 4

I created zones of elevation by quantiles and calculated the average value of landuse diversity in each zone. The trend in the data shows that the fruther you move away from the watershed, the higher the elevation.


```{r message=FALSE, warning=FALSE}

LU_Diversity.rc <- focal(riley_1km, 
                      w = matrix(1, ncol = 3, nrow = 3), 
                      fun = modal, na.rm = T)

#Creating the elevation zones by quantiles
elev_quant <- quantile(values(elev_smooth),
                       probs = c(0, 0.25, 0.50, 0.75, 1.0), 
                       na.rm = T)
elev_zones <- raster(elev_smooth)
values(elev_zones) <- cut(values(elev_smooth),
                          breaks = elev_quant, include.lowest = T)

avg_diversity <- zonal(LU_Diversity.rc,
                       elev_zones, fun = mean)
avg_diversity
plot(elev_zones, main = "Elevation Zones in Riley County")

```



### Question 5

I finally created a raster displaying the distance from developed areas with the closest distance displayed in white and the farthest shown in green. The units are in meters.

```{r message=FALSE, warning=FALSE}
unique(riley_1km)

#reclassify all dev areas to 121
riley_1km <- reclassify(riley_1km, c(120,125, 121))
unique(riley_1km)

##creating developed area points from the raster
dev_area_pts <- rasterToPoints(riley_1km, fun = function(x) {x==121}, spatial = T) 
plot(dev_area_pts)

## creating raster showing distance from developed areas
dev_area_dist <- distanceFromPoints(riley_1km, dev_area_pts) 
plot(dev_area_dist)

```














